---
title: "Patient characteristics tables and analysis for SERTAD2 qPCR"
author: "Sally Beard"
date: "2022-09-28"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This document contains code for generating characteristics tables for qPCRs run for the FOX study, as well as some statistical tests investigating SERTAD2 expression.

```{r, setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r}
#import required libraries
library(tidyverse)
library(tableone)
library(kableExtra)
library(rms)
library(interactionRCS)
```

## Table of characteristics for all TB placenta samples

Import metadata and qPCR results for SERTAD2 and reference genes (geometric mean of CYC1 and TOP1) from placenta RNA

```{r}
metadata <- read.csv("./data/TB_placenta_SERTAD2_metadata.csv")
```

Make a characteristics table

```{r}
#create a variable list
vars <- c("Maternal_age", "BMI", "Diabetes", "Gestation_wks", "Birthweight", "Centile", "Fetal_sex", "Systolic_BP", "Diastolic_BP", "Parity", "Smoking")

#create table stratified by condition
tableOne <- CreateTableOne(vars = vars, strata = "category_3", data = metadata, testNonNormal = kruskal.test)
kable_styling(kableone(tableOne, caption="Table 1: Patient characteristics for samples included in tissue bank placental qPCR across all gestations", nonnormal = c("Birthweight_centile", "Cord_blood_pH", "BMI"), varLabels = TRUE, dropEqual = TRUE))
```
```{r, results = FALSE}
table <- print(tableOne, nonnormal = c("Birthweight_centile", "Cord_blood_pH", "BMI"), exact = "stage", quote = FALSE, noSpaces = TRUE)
write.csv(table, "output/TB_placenta_characteristics.csv")
```

## qPCR results for all TB placenta samples

To test whether SERTAD2 is altered in diseased placenta an additive linear model was applied, which adjusted for gestation using a restricted cubic spline with four knots.

```{r}
model_out <- glm(deltaSERTAD2 ~ rcs(Gestation_days) + category_3, data=metadata)
dd <-datadist(metadata)
options(datadist="dd")
summary(model_out)
```

The expression of SERTAD2 did not significantly differ from control in IUGR alone, PE_IUGR or PE alone placentas across all gestations tested, after adjusting for gestational age. The confidence intervals for the coefficients are shown below.

```{r}
model_out <- ols(deltaSERTAD2 ~ rcs(Gestation_days) + category_3, data=metadata)
summary(model_out, category_3 = "Control")
```

## Table of characteristics for TB placenta samples <34 weeks

Subset the data to just include samples <=34+0 weeks

```{r}
metadata_preterm <- metadata %>% filter(Gestation_wks <= 34)
```

Make a characteristics table

```{r}
#create table stratified by condition
tableOne <- CreateTableOne(vars = vars, strata = "category_3", data = metadata_preterm, testNonNormal = kruskal.test)
kable_styling(kableone(tableOne, caption="Table 2: Patient characteristics for samples included in tissue bank placental qPCR <=34+0 weeks", nonnormal = c("Birthweight_centile", "Cord_blood_pH", "BMI"), varLabels = TRUE, dropEqual = TRUE))
```
```{r, results = FALSE}
table <- print(tableOne, nonnormal = c("Birthweight_centile", "Cord_blood_pH", "BMI"), exact = "stage", quote = FALSE, noSpaces = TRUE)
write.csv(table, "output/TB_placenta_characteristics_preterm.csv")
```

## qPCR results for TB placenta samples <34 weeks

To test whether SERTAD2 is altered in preterm diseased placenta an additive linear model was applied, which adjusted for gestation using a restricted cubic spline with four knots.

```{r}
model_out <- glm(deltaSERTAD2 ~ rcs(Gestation_days) + category_3, data=metadata_preterm)
dd <-datadist(metadata)
options(datadist="dd")
summary(model_out)
```

The expression of SERTAD2 did not significantly differ from control in IUGR alone or PE_IUGR, but PE alone was significantly higher than control placentas, after adjusting for gestational age. The confidence intervals for the coefficients are shown below.

```{r}
model_out <- ols(deltaSERTAD2 ~ rcs(Gestation_days) + category_3, data=metadata)
summary(model_out, category_3 = "Control")
```

Since we identified SERTAD2 in the FOX study from comparing blood from women with preterm growth restricted fetuses with and without preeclampsia, next we tested whether there was a difference in the expression level of SERTAD2 between these two groups.

```{r}
model_out <- ols(deltaSERTAD2 ~ rcs(Gestation_days) + category_3, data=metadata)
summary(model_out, category_3 = "IUGR")
```

The confidence interval for the comparison between PE_IUGR and PE overlaps 0, indicating that there is not a significant difference in SERTAD2 expression between these two groups.

## Table of characteristics for all TB PAX samples

Import metadata and qPCR results for SERTAD2 and reference genes (geometric mean of B2M, GUSB and YWHAZ) from RNA isolated from maternal blood collected into PAXgene tubes.

```{r}
metadata_PAX <- read.csv("./data/TB_PAX_SERTAD2_metadata.csv")
```

Make a characteristics table

```{r}
#create a variable list
vars <- c("Maternal_age", "BMI", "Diabetes", "GA_collection", "GA_delivery", "Birthweight", "Centile", "Fetal_sex", "Systolic_BP", "Diastolic_BP", "Parity", "Smoking")

#create table stratified by condition
tableOne <- CreateTableOne(vars = vars, strata = "Category_2", data = metadata_PAX, testNonNormal = kruskal.test)
kable_styling(kableone(tableOne, caption="Table 3: Patient characteristics for samples included in tissue bank PAX qPCR across all gestations", nonnormal = c( "BMI"), varLabels = TRUE, dropEqual = TRUE))
```
```{r, results = FALSE}
table <- print(tableOne, nonnormal = c( "BMI"), exact = "stage", quote = FALSE, noSpaces = TRUE)
write.csv(table, "output/PAX_characteristics.csv")
```

## qPCR results for all PAX placenta samples

To test whether SERTAD2 is altered in PAX RNA collected from maternal blood, an additive linear model was applied, which adjusted for gestation using a restricted cubic spline with four knots.

```{r}
model_out <- glm(deltaSERTAD2 ~ rcs(GA_collection) + Category_2, data=metadata_PAX)
dd <-datadist(metadata_PAX)
options(datadist="dd")
summary(model_out)
```

The expression of SERTAD2 did not significantly differ from control in IUGR alone, PE_IUGR or PE alone PAX across all gestations tested, after adjusting for gestational age. The confidence intervals for the coefficients are shown below.

```{r}
model_out <- ols(deltaSERTAD2 ~ rcs(GA_collection) + Category_2, data=metadata_PAX)
summary(model_out, Category_2 = "Control")
```